---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit only README.Rmd! -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

# `caracas`: Computer algebra in R

<!-- badges: start -->
[![R build status](https://github.com/r-cas/caracas/workflows/R-CMD-check/badge.svg)](https://github.com/r-cas/caracas/actions) 
[![codecov.io](https://codecov.io/gh/r-cas/caracas/branch/master/graphs/badge.svg)](https://codecov.io/gh/r-cas/caracas?branch=master)
<!-- badges: end -->
  
## Installation

`caracas` is available on CRAN and can be installed as usual:

```
install.packages('caracas')
```

Please ensure that you have SymPy installed, or else install it:

```{r, eval = FALSE}
if (!caracas::has_sympy()) {
  caracas::install_sympy() 
}
```

To build and install from Github with vignettes run this command from within `R` (please install `remotes` first if not already installed):

```
# install.packages('remotes')
remotes::install_github("r-cas/caracas", build_vignettes = TRUE)
```

You can also install the package without vignettes if needed as follows:

```
remotes::install_github("r-cas/caracas")
```

## Configuring the Python environment

The `caracas` package uses the [`reticulate`](https://github.com/rstudio/reticulate) package (to run Python code). Thus, if you wish to configure your Python environment, you need to 1) load `reticulate`, 2) configure the Python environment, and 3) load `caracas`. The Python environment can be configured as described [here](https://rstudio.github.io/reticulate/articles/versions.html). Again, this need to be done *before* loading `caracas`.

## Development site

See <https://github.com/r-cas/caracas>.

## Online documentation

See <https://r-cas.github.io/caracas/>.

## Origin of name

The name "caracas" is intended to mean "(inter)face to computer algebra system(s)" - notice that "cara" is Spanish (Castellano to be precise) for "face".

## Code of conduct

Please note that the `caracas` project is released with a Contributor Code of Conduct (available in `CODE_OF_CONDUCT.md`). By contributing to this project, you agree to abide by its terms.

## Brief introduction

```{r, message=FALSE}
library(caracas)
```

```{r, include = FALSE}
if (!has_sympy()) {
  # SymPy not available, so the chunks shall not be evaluated
  knitr::opts_chunk$set(eval = FALSE)
}
```

```{r}
x <- symbol('x')
eq <- 2*x^2 - x
eq
as.character(eq)
as_expr(eq)
tex(eq)
```

```{r}
solve_sys(eq, x)
der(eq, x)
subs(eq, x, "y")
```

```{r}
A <- matrix(c("x", 2, 0, "2*x"), 2, 2)
B <- as_sym(A)
B
Binv <- inv(B)
Binv
tex(Binv)
```

```{r}
eigenval(Binv)
```

Please find more examples in the other vignettes available at <https://r-cas.github.io/caracas/>.

## Contribute, issues, and support

Please use the issue tracker at <https://github.com/r-cas/caracas/issues> 
if you want to notify us of an issue or need support.
If you want to contribute, please either create an issue or make a pull request.
