---
title: "Chunk engine for easy tex output in Rmarkdown and Quarto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chunk engine for easy tex output in Rmarkdown and Quarto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(caracas)
```


```{r, include = FALSE}
inline_code <- function(x) {
  x
}

if (!has_sympy()) {
  # SymPy not available, so the chunks shall not be evaluated
  knitr::opts_chunk$set(eval = FALSE)
  
  inline_code <- function(x) {
    deparse(substitute(x))
  }
}
```

We have added a chunk engine that outputs (an object) to tex format. 
That is, `tex()` is applied to the object and the output is 
put inside a display math environment.

For example, if you write the following:

````{verbatim}
```{r}
library(caracas)
def_sym(x, y)
p <- x^3 + y^2
```

```{rtex}
der(p, c(x, y))
```
````

You will get this output:

```{r}
library(caracas)
def_sym(x, y)
p <- x^3 + y^2
```

```{rtex}
der(p, c(x, y))
```


# Custom `tex()`

You can also create your own custom `tex()` function:

```{r}
tex <- function(x) {
  paste0("g(x, y) = ", caracas::tex(x))
}
```

```{rtex}
der(p, c(x, y))
```

The reason this works is because `rtex` just calls `tex()` and 
if you define one in global namespace, then this is previous in the 
search path than the one provided by `caracas`.
