---
title: "Concentration and covariance matrix in an autoregressive model and in a dynamic linear model"
author: Mikkel Meyer Andersen and Søren Højsgaard
date: "`r date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Concentration and covariance matrix in an autoregressive model and in a dynamic linear model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE}
library(caracas)
```

```{r, include = FALSE}
inline_code <- function(x) {
  x
}

if (!has_sympy()) {
  # SymPy not available, so the chunks shall not be evaluated
  knitr::opts_chunk$set(eval = FALSE)
  
  inline_code <- inline_code <- function(x) {
    deparse(substitute(x))
  }
}
```

## Autoregressive model ($AR(1)$)

```{r ar1, echo=F}
N <- 3
L1 <- diag(4)
L1[cbind(1 + (1:N), 1:N)] <- "-a"
L1 <- as_sym(L1)
```

```{r, echo=F}
e <- as_sym(paste0("e", 0:3))
x <- as_sym(paste0("x", 0:3))
u <- as_sym(paste0("u", 1:3))
y <- as_sym(paste0("y", 1:3))
eu <- rbind(e, u)
xy <- rbind(x, y)
```


Consider this model: 
$$
x_i = a x_{i-1} + e_i, \quad i=1, \dots, 3
$$ 
and $x_0=e_0$. All terms $e_0, \dots, e_3$ are independent and $N(0,v^2)$ distributed. 
Let $e=(e_0, \dots, e_3)$ and $x=(x_0, \dots x_3)$. Hence $e \sim N(0, v^2 I)$.
Isolating error terms gives
$$
e= `r tex(e)` = `r tex(L1)` `r tex(x)` = L_1 x
$$

```{r}
<<ar1>>
```

Since $\mathbf{Var}(e)=v^2 I$ we have 
$\mathbf{Var}(e)=v^2 I=L \mathbf{Var}(x) L'$ so the covariance matrix of $x$ is $V_1=\mathbf{Var}(x) = v^2 L^- (L^-)'$ 
 while the concentration matrix (the inverse covariances matrix) is $K=v^{-2}L' L$.

```{r}
def_sym(v2)
L1inv <- inv(L1)
V1 <- v2 * L1inv %*% t(L1inv)
K1 <- (t(L1) %*% L1) / v2
```

```{r, echo=F, results="asis"}
cat(
  "\\begin{align} 
    K_1 &= ", tex(K1), " \\\\ 
   V_1 &= ", tex(V1), " 
  \\end{align}", sep = "")
```


## Dynamic linear model

```{r L2, echo=F}
N <- 3
L2chr <- diag("1", 1 + 2*N)
L2chr[cbind(1 + (1:N), 1:N)] <- "-a"
L2chr[cbind(1 + N + (1:N), 1 + 1:N)] <- "-b"
L2 <- as_sym(L2chr)
```

Augment the $AR(1)$ process above with $y_i=b x_i + u_i$ for
$i=1,2,3$.  Suppose $u_i\sim N(0, w^2)$ and all $u_i$ are independent
and inpendent of $e$.
Then
$(e,u)$ can be expressed in terms of $(x,y)$ as
$$
(e,u) = `r tex(eu)` = `r tex(L2)` `r tex(xy)` = L_2 (x,y)
$$
where
```{r}
<<L2>>
```


```{r}
Veu <- diag(1, 7)
diag(Veu)[1:4] <- "v2"
diag(Veu)[5:7] <- "w2"
Veu
Veu <- as_sym(Veu)
Veu
L2inv <- inv(L2) 
V2 <- L2inv %*% Veu %*% t(L2inv) 
K2 <- t(L2) %*% inv(Veu) %*% L2
```

```{r, results="asis", echo=F}
cat(
  "\\begin{align} K_2 &= ", tex(K2), " \\\\ 
                  V_2 &= ", tex(V2), " \\end{align}", sep = "")
```


```{r}
# <!-- ```{r} -->
# <!-- x_index <- 1:4 -->
# <!-- Kxx <- K2[x_index, x_index] -->
# <!-- Kxy <- K2[x_index, -x_index] -->
# <!-- Kyx <- K2[-x_index, x_index] -->
# <!-- Kyy <- K2[-x_index, -x_index] -->
# 
# <!-- S <- Kyy - Kyx %*% inv(Kxx) %*% Kxy -->
# <!-- Si <- inv(S) -->
# <!-- ``` -->
# 
# <!-- $$ -->
# <!--  S^{-1} = `r tex(Si)` -->
# <!-- $$ -->
#    
#    
#    
#    
# 
# <!-- ## Numerical evalation in R -->
# 
# <!-- ```{r} -->
# <!-- sparsify <- function(x) { -->
# <!--   if (requireNamespace("Matrix", quietly = TRUE)) { -->
# <!--       return(Matrix::Matrix(x, sparse = TRUE)) -->
# <!--       ##return(Matrix(x, sparse = TRUE)) -->
# <!--   } -->
#   
# <!--   return(x) -->
# <!-- } -->
# 
# <!-- alpha <- 0.5 -->
# <!-- beta <- -0.3 -->
# 
# <!-- ## AR(1) -->
# <!-- N <- 3 -->
# <!-- L1 <- diag(1, 1 + N) -->
# <!-- L1[cbind(1+(1:N), 1:N)] <- -alpha -->
# <!-- K1 <- L1 %*% t(L1) -->
# <!-- V1 <- solve(K1) -->
# <!-- sparsify(K1) -->
# <!-- sparsify(V1) -->
# 
# <!-- ## Dynamic linear models -->
# <!-- N <- 3 -->
# <!-- L2 <- diag(1, 1 + 2*N) -->
# <!-- L2[cbind(1+(1:N), 1:N)] <- -alpha -->
# <!-- L2[cbind(1 + N + (1:N), 1 + 1:N)] <- -beta -->
# <!-- K2 <- L2 %*% t(L2) -->
# <!-- V2 <- solve(K2) -->
# <!-- sparsify(K2) -->
# <!-- sparsify(V2) -->
# <!-- ``` -->
# 
# <!-- Comparing with results calculated by `caracas`: -->
# 
# <!-- ```{r} -->
# <!-- V1_eval <- eval(as_expr(V1), list(a = alpha)) -->
# <!-- V2_eval <- eval(as_expr(V2, first_doit = FALSE), list(a = alpha, b = beta)) -->
# <!-- all.equal(V1, V1_eval) -->
# <!-- all.equal(V2, V2_eval) -->
# <!-- ```  -->
```




