---
title: "Fastest route through the forest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastest-route}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE}
library(caracas)
```

```{r, include = FALSE}
if (!have_sympy()) {
  # SymPy not available, so the chunks shall not be evaluated
  knitr::opts_chunk$set(eval = FALSE)
}
```

A [problem](https://ing.dk/artikel/taenkeboks-hurtigste-vej-mellem-a-b-1150-m-lang-232871) was posted by the Danish newspaper, *Ingeniøren*, and it goes like this:

You are in a forest located at $A$. 
You need to get to $B$ in the fastest way possible. 
But you can only change direction once. 
You can take the longer walking path ($AC$ and $CB$) where you can walk fast, 
the direct path through the forest ($AB$) where you have to walk slower, 
or first take the fast walking path some distance, $k$, and then cross through the forest ($AD$, $DB$).

```{r, echo = FALSE, fig.width=8, fig.height=5}
old_mar <- par("mar")
par(mar = c(0, 0, 0, 0))
A <- 300
kat <- sqrt(1000^2 - A^2)

pA <- c(0, 0)
pB <- c(kat, A)
pC <- c(kat, 0)

D <- 400
pD <- c(D, 0)

plot(0, type = 'n', axes = FALSE, ann = FALSE, xlim = c(0, 1075), ylim = c(-50, 350))

polygon(c(pA[1], pB[1], pC[1]), c(pA[2], pB[2], pC[2]), border = NA, col = "green")

points(pA[1], pA[2], type = "p"); text(pA[1], pA[2], labels = "A", pos = 2)
points(pB[1], pB[2], type = "p"); text(pB[1], pB[2], labels = "B", pos = 3)
points(pC[1], pC[2], type = "p"); text(pC[1], pC[2], labels = "C", pos = 4)
points(pD[1], pD[2], type = "p"); text(pD[1], pD[2], labels = "D", pos = 1)

lines(c(pA[1], pB[1]), c(pA[2], pB[2]), type = "l", col = "green")
lines(c(pA[1], pC[1]), c(pA[2], pC[2]), type = "l", col = "black")
lines(c(pB[1], pC[1]), c(pB[2], pC[2]), type = "l", col = "black")
lines(c(pD[1], pB[1]), c(pD[2], pB[2]), type = "l", col = "black", lty = 2)

text(kat/5, 0, labels = "v1 = 5 m/s", pos = 1)
text(kat, A/2, labels = "d1 = 300 m\nv1 = 5 m/s", pos = 4)

text(kat/2, A/1.8, labels = "d2 = 1,000 m\nv2 = 2 m/s", pos = 3)

par(mar = old_mar)
```

## Information given

* Distances
  * $|CB| =$ 300 m
  * $|AB| =$ 1,000 m
* Velocities
  * $v_{AC} = 5$ m/s
  * $v_{CB} = 5$ m/s
  * $v_{AB} = 2$ m/s
  * $v_{DB} = 2$ m/s

### Length of line segments

We parameterise with $k = |AD|$, the distance between $A$ and $D$. 
That is, how much to walk on fast walking path before crossing into 
the forest.

Formulating using `caracas`:

```{r}
CB <- as_symbol('300')
CB
AB <- as_symbol('1000')
AB
AC <- sqrt(AB^2 - CB^2)
AC
k <- symbol('k') # AD
DC <- AC - k
DB <- sqrt(DC^2 + CB^2)
DB
```

So for a distance of $k$, you travel by 5 m/s, and then for a distance of 
$`r tex(DB)`$ you travel by 2 m/s. 
Thus it takes $k/5$ to travel $AD$ and $`r tex(DB/2)`$ to travel $DB$.

At the two extreme values for $k = 0$ and $k = |AC|$ we get:

```{r}
# k = 0: route is AB
as_r(AB)/2
# k = |AC|: route is AC + CB
(as_r(AC) + as_r(CB))/5
```

The question is: Can we do better?

First, the total duration of the route is:

```{r}
l <- k/5 + DB/2
l
```

(The discontinuity at $k = |AC|$ is ignored.)

```{r, fig.width=4, fig.height=4}
lfun <- as_r(l)
lfun
ks <- seq(0, as_r(AC), length.out = 100)
ls <- eval(lfun, list(k = ks))
plot(ks, ls, type = "l")
```

It looks like a minimum around $k = 800$.

We find the analytical solution by first finding critical points:

```{r}
dl <- dd(l, k)
dl
crit_points <- rootsolve(dl, k)
crit_points
best_k <- crit_points[[1]]$k
best_k
```

The type of the critical point is found by considering the Hessian:

```{r}
eval(as_r(dd(dl, k)), list(k = as_r(best_k)))
```

Thus the critical point is indeed a minimum as suggested by the plot.

The fastest route is thus obtained for 
\[
  k = `r tex(best_k)` \approx `r round(as_r(best_k), 2)` .
\]
It has a length of (in meters)
```{r}
DC_best <- AC - best_k
DB_best <- sqrt(DC_best^2 + CB^2)
DB_best
best_route <- best_k + DB_best
best_route
as_r(best_route)
```
\[
  `r tex(best_route)` \approx `r round(as_r(best_route), 2)`
\]
and takes (in seconds)
```{r}
best_l <- subs(l, "k", best_k)
best_l
as_r(best_l)
```
\[
  `r tex(best_l)` \approx `r round(as_r(best_l), 2)`
\]

The best route can be illustrated, too:

```{r, echo = FALSE, fig.width=8, fig.height=5}
old_mar <- par("mar")
par(mar = c(0, 0, 0, 0))
A <- 300
kat <- sqrt(1000^2 - A^2)

pA <- c(0, 0)
pB <- c(kat, A)
pC <- c(kat, 0)

D <- as_r(best_k)
pD <- c(D, 0)

plot(0, type = 'n', axes = FALSE, ann = FALSE, xlim = c(0, 1075), ylim = c(-50, 350))

polygon(c(pA[1], pB[1], pC[1]), c(pA[2], pB[2], pC[2]), border = NA, col = "green")

points(pA[1], pA[2], type = "p"); text(pA[1], pA[2], labels = "A", pos = 2)
points(pB[1], pB[2], type = "p"); text(pB[1], pB[2], labels = "B", pos = 3)
points(pC[1], pC[2], type = "p"); text(pC[1], pC[2], labels = "C", pos = 4)
points(pD[1], pD[2], type = "p"); text(pD[1], pD[2], labels = "D", pos = 1)

lines(c(pA[1], pB[1]), c(pA[2], pB[2]), type = "l", col = "green")
lines(c(pA[1], pC[1]), c(pA[2], pC[2]), type = "l", col = "black")
lines(c(pB[1], pC[1]), c(pB[2], pC[2]), type = "l", col = "black")

lines(c(pA[1], pD[1]), c(pA[2], pD[2]), type = "l", col = "red", lwd = 2)
lines(c(pD[1], pB[1]), c(pD[2], pB[2]), type = "l", col = "red", lwd = 2)

text(kat/5, 0, labels = "v1 = 5 m/s", pos = 1)
text(kat, A/2, labels = "d1 = 300 m\nv1 = 5 m/s", pos = 4)

text(kat/2, A/1.8, labels = "d2 = 1,000 m\nv2 = 2 m/s", pos = 3)

par(mar = old_mar)
```

We end up concluding that the above is the best route if forced to 
go through forest. 
But if one can choose to use the dedicated walking path the entire way, 
then that is actually the fastest. 
In contrast to what the [answer](https://ing.dk/artikel/taenkeboks-hurtigste-vej-mellem-a-b-1150-m-lang-232871) originally says.



