---
title: "Fastest route through the forest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fastest route through the forest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE}
library(caracas)
```

```{r, include = FALSE}
inline_code <- function(x) {
  x
}

if (!has_sympy()) {
  # SymPy not available, so the chunks shall not be evaluated
  knitr::opts_chunk$set(eval = FALSE)
  
  inline_code <- inline_code <- function(x) {
    deparse(substitute(x))
  }
}
```

A [problem](https://ing.dk/artikel/taenkeboks-hurtigste-vej-mellem-a-b-1150-m-lang-232871) was posted by the Danish newspaper, *Ingeniøren*, and it goes like this:

You are in the middle of a dense forest located at $A$. 
You need to get to $C$ in the fastest way possible, and you can only change direction once. 
You can walk directly via $AB$ to the dedicated walking path $BC$ where you can walk fast, 
you can take the direct path through the forest ($AC$) where you have to walk slower, 
or cross through the forest to the dedicated walking path ($AD$ and then $DC$).

```{r, echo = FALSE, fig.width=8, fig.height=5}
old_mar <- par("mar")
par(mar = c(0, 0, 0, 0))
A <- 300
kat <- sqrt(1000^2 - A^2)

pA <- c(0, A)
pB <- c(0, 0)
pC <- c(kat, 0)

D <- 400
pD <- c(D, 0)

pad <- 40
plot(0, type = 'n', axes = FALSE, ann = FALSE, 
     xlim = c(-pad, 1000+pad), 
     ylim = c(-pad, A+pad))

polygon(c(pA[1]-pad, pB[1]-pad, pC[1]+pad, pC[1]+pad), 
        c(pA[2]+pad, pB[2]-pad, pC[2]-pad, pA[2]+pad), border = NA, col = "green")

points(pA[1], pA[2], type = "p"); text(pA[1], pA[2], labels = "A", pos = 2)
points(pB[1], pB[2], type = "p"); text(pB[1], pB[2], labels = "B", pos = 2)
points(pC[1], pC[2], type = "p"); text(pC[1], pC[2], labels = "C", pos = 4)
points(pD[1], pD[2], type = "p"); text(pD[1], pD[2], labels = "D", pos = 1)

lines(c(pA[1], pB[1]), c(pA[2], pB[2]), type = "l", col = "black", lty = 2)
lines(c(pA[1], pC[1]), c(pA[2], pC[2]), type = "l", col = "black", lty = 2)
lines(c(pB[1], pC[1]), c(pB[2], pC[2]), type = "l", col = "black", lwd = 3)
lines(c(pA[1], pD[1]), c(pA[2], pD[2]), type = "l", col = "black", lty = 2)

legend(1000-200, A, 
       legend = c("5 m/s", "2 m/s"), 
       lty = c(1, 2), 
       lwd = c(3, 1))
text(-pad/3, A/2, labels = "|AB| = 300 m", pos = 3, srt = 90)
text(kat/2, A/1.7, labels = "|AC| = 1,000 m", pos = 3)

par(mar = old_mar)
```

## Information given

* Distances
  * $|AB| =$ 300 m
  * $|AC| =$ 1,000 m
* Velocities
  * $v_{AB} = 2$ m/s
  * $v_{AD} = 2$ m/s
  * $v_{AC} = 2$ m/s
  * $v_{BC} = 5$ m/s

### Length of line segments

We parameterise with $k = |BD|$, the distance between $B$ and $D$. 
That is, how much to walk on fast walking path before crossing into 
the forest.

Formulating using `caracas`:

```{r}
AB <- as_sym('300')
AB
AC <- as_sym('1000')
AC
BC <- sqrt(AC^2 - AB^2)
BC
k <- symbol('k') # |BD|
DC <- BC - k
AD <- sqrt(AB^2 + k^2)
AD
```

So for a distance of $|AD|$, you travel by 5 m/s, and then for a distance of 
$`r inline_code(tex(DC))`$ you travel by 2 m/s. 
Thus it takes $`r inline_code(tex(AD/2))`$ to travel $AD$ and $`r inline_code(tex(DC/5))`$ to travel $DC$.

The question is: What is the fastest way to get from $A$ to $C$?

First, the total duration of the route is:

```{r}
l <- AD/2 + DC/5
l
```

```{r, fig.width=7, fig.height=5}
lfun <- as_expr(l)
lfun
ks <- seq(0, as_expr(AC), length.out = 100)
ls <- eval(lfun, list(k = ks))
plot(ks, ls, type = "l", xlab = "k", ylab = "Time A to C")
```

It looks like a minimum around $k = 150$.

We find the analytical solution by first finding critical points:

```{r}
dl <- der(l, k)
dl
crit_points <- solve_sys(dl, k)
crit_points
best_k <- crit_points[[1]]$k
best_k
```

The type of the critical point is found by considering the Hessian:

```{r}
eval(as_expr(der(dl, k)), list(k = as_expr(best_k)))
```

Thus the critical point is indeed a minimum as suggested by the plot.

The fastest route is thus obtained for 
\[
  k = `r inline_code(tex(best_k))` \approx `r inline_code(round(as_expr(best_k), 2))` .
\]
It has a length of (in meters)
```{r}
DC_best <- BC - best_k
AD_best <- sqrt(AB^2 + best_k^2)
AD_best
best_route <- AD_best + DC_best
best_route
as_expr(best_route)
```
\[
  `r inline_code(tex(best_route))` \approx `r inline_code(round(as_expr(best_route), 2))`
\]
and takes (in seconds)
```{r}
best_l <- subs(l, "k", best_k)
best_l
as_expr(best_l)
```
\[
  `r inline_code(tex(best_l))` \approx `r inline_code(round(as_expr(best_l), 2))`
\]

The best route can be illustrated, too:


```{r, echo = FALSE, fig.width=8, fig.height=5}
old_mar <- par("mar")
par(mar = c(0, 0, 0, 0))
A <- 300
kat <- sqrt(1000^2 - A^2)

pA <- c(0, A)
pB <- c(0, 0)
pC <- c(kat, 0)

D <- as_expr(best_k)
pD <- c(D, 0)

pad <- 40
plot(0, type = 'n', axes = FALSE, ann = FALSE, 
     xlim = c(-pad, 1000+pad), 
     ylim = c(-pad, A+pad))

polygon(c(pA[1]-pad, pB[1]-pad, pC[1]+pad, pC[1]+pad), 
        c(pA[2]+pad, pB[2]-pad, pC[2]-pad, pA[2]+pad), border = NA, col = "green")

points(pA[1], pA[2], type = "p"); text(pA[1], pA[2], labels = "A", pos = 2)
points(pB[1], pB[2], type = "p"); text(pB[1], pB[2], labels = "B", pos = 2)
points(pC[1], pC[2], type = "p"); text(pC[1], pC[2], labels = "C", pos = 4)
points(pD[1], pD[2], type = "p"); text(pD[1], pD[2], labels = "D", pos = 1)

lines(c(pA[1], pB[1]), c(pA[2], pB[2]), type = "l", col = "black", lty = 2)
lines(c(pA[1], pC[1]), c(pA[2], pC[2]), type = "l", col = "black", lty = 2)
lines(c(pB[1], pC[1]), c(pB[2], pC[2]), type = "l", col = "black", lwd = 3)
lines(c(pA[1], pD[1]), c(pA[2], pD[2]), type = "l", col = "black", lty = 2)

lines(c(pA[1], pD[1]), c(pA[2], pD[2]), type = "l", col = "red", lwd = 2.5)
lines(c(pD[1], pC[1]), c(pD[2], pC[2]), type = "l", col = "red", lwd = 2.5)

legend(1000-200, A, 
       legend = c("5 m/s", "2 m/s"), 
       lty = c(1, 2), 
       lwd = c(3, 1))
text(-pad/3, A/2, labels = "|AB| = 300 m", pos = 3, srt = 90)
text(kat/2, A/1.7, labels = "|AC| = 1,000 m", pos = 3)

par(mar = old_mar)
```
