---
title: "Introduction to 'caracas'"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to 'caracas'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(caracas)
```

```{r, include = FALSE}
if (!have_sympy()) {
  # SymPy not available, so the chunks shall not be evaluated
  knitr::opts_chunk$set(eval = FALSE)
}
```

## Road-map

Currently only a `SymPy` object is available. 
More functionality will be added shortly. 
Please keep an eye out at the development site at <https://github.com/r-cas/caracas>.

## Notes

* Python syntax is used, which means that there are a few differences to R's syntax:
  + Power is `**` not `^` (see below)

## Quick start

```{r}
x <- symbol('x')
as.character(x)
x
as_r(x)
```


```{r}
2*x
y <- symbol('y')
sqrt(3*x^y)
```

```{r}
z <- cos(x)^2 + sin(x)^2
z
simplify(z)
tex(z)
```

```{r}
z <- cos(x)*cos(y) - sin(x)*sin(y)
z
simplify(z)
z <- cos(x + y)
z
expand(z)
expand_trig(z)
```

```{r}
x <- symbol('x')
y <- symbol('y')
z <- log(x*y)
z
expand_log(z)
```


### Sums

```{r}
x <- symbol("x")
sumf(1/x, "x", 1, 10)
sumf(1/x, x, 1, 10)
s <- sumf(1/x, "x", 1, 10)
as_r(s)
sum(1/(1:10))
n <- symbol("n")
simplify(sumf(x, x, 1, n))
```


### Products

```{r}
x <- symbol("x")
p <- prodf(1/x, "x", 1, 10)
p
as_r(p)
prod(1/(1:10))
n <- symbol("n")
prodf(x, x, 1, n)
```


### Integrals

```{r}
x <- symbol("x")

intf(1/x, x, 1, 10)
i1 <- intf(1/x, x, 1, 10, doit = FALSE)
i1
tex(i1)
doit(i1)
intf(1/x, x)
i1 <- intf(1/x, x, doit = FALSE)
i1
tex(i1)
doit(i1)
```

### Limits


```{r}
x <- symbol("x")
lim(sin(x)/x, "x", 0)
lim(1/x, "x", 0, dir = '+')
lim(1/x, "x", 0, dir = '-')
```

We can also postpone evaluation:

```{r}
x <- symbol("x")
lim(sin(x)/x, "x", 0)
lim(sin(x)/x, x, 0)
```

```{r}
res <- lim(sin(x)/x, "x", 0, doit = FALSE)
res
as.character(res)
tex(res)
doit(res)
as_r(res)
```

### Derivatives

Note that the function is called `d()` and not `deriv()`.

```{r}
x <- symbol("x")
y <- symbol("y")
f <- 3*x^2 + x*y^2
f
as_r(f)
dd(f, "x")
dd(f, x)
dd(f, c("x", "y"))
dd(f, c(x, y))
f1 <- dd(f, c(x, y))
f1
as.character(f1)
as_r(f1)
eval(as_r(f1), list(x = 1, y = 2))
dd(f1, c(x, y))
f2 <- dd2(f, c(x, y))
f2
as_r(f2)
eval(as_r(f2), list(x = 1, y = 2))
```


```{r}
x <- symbol("x")
y <- symbol("y")
f <- eval_to_symbol("[3*x**2 + x*y**2, 2*x, 5*y]")
f
dd(f, c(x, y))
```


## Linear algebra

```{r}
A <- matrix(c("x", 0, 0, "2*x"), 2, 2)
A
B <- as_symbol(A)
B
2*B
dim(B)
sqrt(B)
log(B)
B %*% t(B)
```

```{r}
A <- matrix(c("a", 0, 0, 0, "a", "a", "a", 0, 0), 3, 3)
B <- as_symbol(A)
eigenvals(B)
eigenvects(B)
eigen(eval(as_r(B), list(a = 2)))
```



## Solve

* `inverse()`
* `linsolve()`
* `rootsolve()`
* `systemsolve()`

```{r}
# Multinomial likelihood
p <- as_symbol(paste0("p", 1:3))
y <- as_symbol(paste0("y", 1:3))
a <- as_symbol("a")
l <- sum(y*log(p))
l
L <- -l + a*(sum(p) - 1)
L
tex(L)
g <- dd(L, c("a", paste0("p", 1:3)))
g
sol <- rootsolve(g, c("p1", "p2", "p3", "a"))
sol
tex(sol[[1L]]$p1)
```

## Subsetting

```{r}
A <- matrix(c("a", 0, 0, 0, "a", "a", "a", 0, 0), 3, 3)
B <- as_symbol(A)
B[, 2]
B[, -2]
B[1, ]
B[, 2] <- "x"
B
```


## Using `SymPy` directly

```{r}
sympy <- get_sympy()
```

```{r}
sympy$diff("2*a*x", "x")
sympy$solve("x**2 - 1", "x")
```

## More examples

Please find more examples in the other vignettes.

